---
title: "Lab 1"
author: "Roxanne Li"
date: "2024-01-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Read in some packages that we'll be using:

```{r}
#| message: false
#install.packages("tidyverse")
library(tidyverse)
```

Read in mortality rates for Ontario:
```{r}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm)
```
## Q1

Plot the ratio of female to male mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme.

```{r}
dm1 <- dm |> 
  filter(Age==10|Age==20|Age==30|Age==40) |>
  mutate(mf_ratio = Female/Male) 
```

```{r}
dm1 |> 
  ggplot(aes(x = Year, y = mf_ratio, color = Age)) + 
  geom_line() + 
  scale_color_brewer(palette = "Set1") +
  labs(title = "Ratio of Female to Male Mortality Rates Over Time",
       x = "Year",
       y = "Ratio of Female to Male Mortality Rates") +
  theme_minimal() 
```

## Q2

Find the age that has the lowest female mortality rate each year.

```{r}
dm2 <- dm |> 
  group_by(Year) |> 
  slice(which.min(Female)) |>
  select(Year, Age, Female)
dm2
```
The table produced above shows the age that has the lowest female mortality rate each year.

## Q3

Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations. 

```{r}
dm |> 
  group_by(Age) |> 
  summarize(across(c(Male, Female, Total), sd, na.rm = TRUE))
```


## Q4

The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here. 

Reading the population file:
```{r}
pop <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", 
                  skip = 2, col_types = "dcddd")
head(pop)
```

Left joining the mortality rate data with the population data and calculating population-weighted mortality rates for females and males:

```{r}
merged_data <- left_join(dm, pop, by=c("Year", "Age"), suffix=c(".dm", ".pop"), unmatched = "drop")
merged_data$Pop_Weighted_Female <- (merged_data$Female.dm + merged_data$Male.dm) * 
  (merged_data$Female.pop / merged_data$Total.pop)
merged_data$Pop_Weighted_Male <- (merged_data$Female.dm + merged_data$Male.dm) * 
  (merged_data$Male.pop / merged_data$Total.pop)
head(merged_data)
```

Calculating population-weighted average mortality rates for females and males, by year.

```{r}
pop_weighted_avg <- merged_data |> 
  group_by(Year) |> 
  summarize(across(c(Pop_Weighted_Female, Pop_Weighted_Male), mean, na.rm = TRUE))
head(pop_weighted_avg)
```
Turning the `pop_weighted_avg` into a pivot table for plotting:

```{r}
pop_weighted_avg_pivot <- pop_weighted_avg |> 
  pivot_longer(Pop_Weighted_Female:Pop_Weighted_Male, names_to = "Sex", 
               values_to = "Population_Weighted_Mortality")
head(pop_weighted_avg_pivot)
```
Plotting the line plot for the population weighted average mortality rates of females and males over time:

```{r}
pop_weighted_avg_pivot |> 
  ggplot(aes(x = Year, y = Population_Weighted_Mortality, color = Sex)) + 
  geom_line() + 
  scale_color_brewer(palette = "Set1") +
  labs(title = "Population-Weighted Average Mortality Rates of Females and Males Over Time",
       x = "Year",
       y = "Mortality") +
  theme_minimal() 
```
The population weighted average mortality rates show that over time, the rates for females have an upward trend and the rates for males have a downward trend.


## Q5

Write down using appropriate notation, and run a simple linear regression with logged mortality rates as the outcome and age (as a continuous variable) as the covariate, using data for females aged less than 106 for the year 2000. Interpret the coefficient on age. 

Getting the data:
```{r}
data <- dm |> 
  filter(Year==2000,  Age<106) |>
  select(c(Year, Age, Female))
head(data)
```
We fit a linear regression model $log(y_i) = \beta_{0} + \beta_{1}x_i + \epsilon_i$ where $y_i$ represents the female mortality rate, $x_i$ represents the female age, and $\epsilon_i$ represents the random error.

```{r}
data$Age <- as.numeric(data$Age)
model <- lm(log(Female) ~ Age, data)
summary(model)
```
The coefficient for `Age` is approximately 0.07. This means on average, if we keep everything else constant, then a female being one year older will be $e^{0.07} \approx 1.07$ times more likely to die in the year of 2000.

