---
title: "Lab 10"
author: "Roxanne Li"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Child mortality in Sri Lanka

In this lab you will be fitting a couple of different models to the data about child mortality in Sri Lanka, which was used in the lecture. Here's the data and the plot from the lecture:

```{r}
library(tidyverse)
library(here)
library(rstan)
library(tidybayes)

lka <- read_csv("data/lka.csv")
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

# Fitting a linear model 

Let's firstly fit a linear model in time to these data. Here's the code to do this:

```{r}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(y = lka$logit_ratio, 
                  year_i = observed_years - years[1]+1, 
                  T = nyears, 
                  years = years, 
                  N = length(observed_years), 
                  mid_year = mean(years), 
                  se = lka$se)

mod <- stan(data = stan_data,
             file = "code/models/lka_linear_me.stan")

```

Extract the results:

```{r}
res <- mod %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year = years[t])
```


Plot the results:

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res, aes(year, .value)) + 
  geom_ribbon(data = res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Linear fit shown in black")
```

## Question 1

Project the linear model above out to 2022 by adding a `generated quantities` block in Stan (do the projections based on the expected value $\mu$). Plot the resulting projections on a graph similar to that above. 

Extract future predictions and plot the results:

```{r}
pred <- mod %>% 
  gather_draws(future_mu[t]) %>% 
  median_qi() %>% 
  mutate(year = max(years)+t)

ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw() +
  geom_line(data = res, aes(year, .value)) + 
  geom_line(data = pred, aes(year, .value), col="blue") +
  geom_ribbon(data = res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_ribbon(data = pred, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill="blue") +
  theme_bw() +
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Linear fit shown in black; Projections shown in blue")

```

## Question 2

The projections above are for the logit of the ratio of neonatal to under-five child mortality. You can download estimates of the under-five child mortality from 1951 to 2022 here: https://childmortality.org/all-cause-mortality/data/estimates?refArea=LKA. Use these data to get estimates and projections of neonatal mortality for Sri Lanka, and plot the results. 

Loading the estimates of the under-five child mortality from 1951 to 2022:

```{r}
under_five <- read.csv("data/lka_under_five.csv")
```

Merge the under five mortality estimates with the logit ratios:

```{r}
bind <- rbind(data.frame(Year = res$year, lratio = res$.value, ll = res$.lower, lu = res$.upper),
              data.frame(Year = pred$year, lratio = pred$.value, ll = pred$.lower, lu = pred$.upper))
bind$ratio <- plogis(bind$lratio)
bind$ratio_lower <- plogis(bind$ll)
bind$ratio_upper <- plogis(bind$lu)

bind <- merge(bind, under_five, by = "Year", all.x = TRUE)
```

Calculate estimates and projections as well as the lower bounds and upper bounds of neonatal mortality:

```{r}
bind$neo <- bind$ratio * bind$Estimate / 1000
bind$neo_lower <- bind$ratio_lower * bind$Lower.bound/ 1000
bind$neo_upper <- bind$ratio_upper * bind$Upper.bound / 1000
```

Plot the results:

```{r}
bind_estimates <- bind |>
  filter(Year <= max(years))

bind_projections <- bind |>
  filter(Year > max(years))

ggplot(data = bind, aes(Year, neo)) +
  geom_line(data = bind_estimates, aes(Year, neo)) +
  geom_line(data = bind_projections, aes(Year, neo), col="blue") +
  geom_ribbon(data = bind_estimates, aes(y = neo, ymin = neo_lower, ymax = neo_upper), alpha = 0.2) +
  geom_ribbon(data = bind_projections, aes(y = neo, ymin = neo_lower, ymax = neo_upper), alpha = 0.2, fill="blue") +
  theme_bw() +
  labs(title = "Neonatal mortality for Sri Lanka",
       y = "Neonatal mortality", subtitle = "Estimates shown in black; Projections shown in blue")
```


# Random walks


## Question 3

Code up and estimate a first order random walk model to fit to the Sri Lankan data, taking into account measurement error, and project out to 2022. 

```{r}
stan_data3 <- list(N = length(observed_years),
                  T = nyears, 
                  P = 8,
                  mid_year = mean(years),
                  y = lka$logit_ratio, 
                  se = lka$se,
                  years = years,
                  year_i = observed_years - years[1]+1
                  )

mod3 <- stan(data = stan_data3,
             file = "code/models/lka_Q3.stan",
             seed = 1234)
```

Extract and plot the results:

```{r}
res3 <- mod3 |>
  gather_draws(mu[t]) |> 
  median_qi() |> 
  mutate(year = years[t])

pred3 <- mod3 |> 
  gather_draws(future_mu[p]) |> 
  median_qi() |> 
  mutate(year = max(years)+p)

ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res3, aes(year, .value)) + 
  geom_line(data = pred3, aes(year, .value), col="blue") +
  geom_ribbon(data = res3, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_ribbon(data = pred3, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill="blue") +
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Estimates shown in black; Projections shown in blue")
```


## Question 4

Now alter your model above to estimate and project a second-order random walk model (RW2). 

```{r}
stan_data4 <- list(N = length(observed_years),
                  T = nyears, 
                  P = 8,
                  mid_year = mean(years),
                  y = lka$logit_ratio, 
                  se = lka$se,
                  years = years,
                  year_i = observed_years - years[1]+1
                  )

mod4 <- stan(data = stan_data4,
             file = "code/models/lka_Q4.stan",
             seed = 1234)
```

Extract and plot the results:

```{r}
res4 <- mod4 |>
  gather_draws(mu[t]) |> 
  median_qi() |> 
  mutate(year = years[t])

pred4 <- mod4 |> 
  gather_draws(future_mu[p]) |> 
  median_qi() |> 
  mutate(year = max(years)+p)

ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res4, aes(year, .value)) + 
  geom_line(data = pred4, aes(year, .value), col="blue") +
  geom_ribbon(data = res4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_ribbon(data = pred4, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill="blue") +
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Estimates shown in black; Projections shown in blue")
```

## Question 5

Run the first order and second order random walk models, including projections out to 2022. Compare these estimates with the linear fit by plotting everything on the same graph. 

```{r}
linear_fit <- rbind(res[, c(".value", ".lower",".upper","year")],
                    pred[, c(".value", ".lower",".upper","year")])
linear_fit$type <- "linear_fit"

rw1 <- rbind(res3[, c(".value", ".lower",".upper","year")],
             pred3[, c(".value", ".lower",".upper","year")])
rw1$type <- "RW1"

rw2 <- rbind(res4[, c(".value", ".lower",".upper","year")],
             pred4[, c(".value", ".lower",".upper","year")])
rw2$type <- "RW2"

plot_data <- rbind(linear_fit, rw1, rw2)

ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes(color = source)) + 
  geom_line(aes(color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill = source), alpha = 0.1) + 
  theme_bw() +
  geom_line(data = plot_data, aes(year, .value, color = type)) +
  geom_ribbon(data = plot_data, aes(y = .value, ymin = .lower, ymax = .upper, fill = type), alpha = 0.2) +
  theme_bw() +
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio",
       subtitle = "Observed value shown in dashed line; Estimates and projections shown in line")
```


## Question 6

Briefly comment on which model you think is most appropriate, or an alternative model that would be more appropriate in this context. 

I think the first order random walk model is most appropriate. From the plot in question 5, we can see that the linear model underfits the data as it has large bias and small variance, and the second order random walk model overfits the data as it has small bias and large variance. The first order random walk model captures a more precise trend of the data than the linear model, and the manner in which it does so is not as oscillating as the second order model. 

